---
title: "03_modeling"
author: "Maksym Khavil"
date: "2025-07-19"
output: html_document
---

## Setup 

Let us load the data and libraries.
```{r}
library(stats)
library(glmnet)
library(rpart)
library(rpart.plot)
library(pROC)
library(corrplot)
library(car)

source(here::here("R", "constants.R"))
source(here::here("R", "load_data.R"))

data <- load_bank_data()
data
```

We will also define a generic function to run cross-validation. That way we will estimate accuracy on new data, we also collect all predictions to construct confusion matrix and Receiver Operator Curve. We will display AUC ROC, accuracy, f1 score, precision and recall. We model data to better profile, so we choose the metric of performance to be precision, as it is a measure of confidence of model. Nevertheless precision can be abused, so we will also keep an eye on accuracy and f1 score.

```{r}
cv_evaluate_model <- function(model_func, predict_func, weights = NULL, k = 5, seed = 42, arg_data=data, ...) {
      set.seed(seed)
    
      n <- nrow(data)
      folds <- sample(rep(1:k, length.out = n))
      
      y_true_all <- c()
      y_pred_all <- c()
      y_prob_all <- c()
    
      for (i in 1:k) {
          train_idx <- which(folds != i)
          test_idx  <- which(folds == i)
      
          train_data <- arg_data[train_idx, ]
          test_data  <- arg_data[test_idx, ]
          model <- NULL
          if(!is.null(weights)){
            train_weights <- weights[train_idx]
            model <- model_func(y ~ ., data=train_data, weights=train_weights, ...)
          }
          else {
            model <- model_func(y ~ ., data=train_data, ...)
          }
      
          
          prob <- predict_func(model, test_data)
          pred <- ifelse(prob > 0.5, "yes", "no")
          
          y_true_all <- c(y_true_all, test_data[['y']])
          y_pred_all <- c(y_pred_all, pred)
          y_prob_all <- c(y_prob_all, prob)
      }
    
      y_true_all <- factor(ifelse(y_true_all == 1, 'no', 'yes'), levels = c("no", "yes"))
      y_pred_all <- factor(y_pred_all, levels = c("no", "yes"))
    
      cm <- table(Predicted = y_pred_all, Actual = y_true_all)
    
      tp <- cm["yes", "yes"]
      tn <- cm["no", "no"]
      fp <- cm["yes", "no"]
      fn <- cm["no", "yes"]
    
      accuracy <- (tp + tn) / sum(cm)
      precision <- tp / (tp + fp)
      recall <- tp / (tp + fn)
      f1 <- 2 * precision * recall / (precision + recall)
    
      
      corrplot(
          cm,
          is.corr = F, method = "color", tl.col = "black", addCoef.col = "black",
          xlab = "Ground Truth", ylab = "Prediction", mar=c(3,0,3,0),
          title = "Confusion Matrix", cl.pos = 'n', col=c('bisque', 'lightblue')
      )
      mtext("Ground Truth", side = 3, line=1, cex = 1.2)
      mtext("Predictions", side = 2, line=0, cex = 1.2)
    
      roc_obj <- roc(y_true_all, y_prob_all, levels=c("no", "yes"), direction='<')
      plot(roc_obj, col = "red", main = "ROC Curve")
      
      legend(
          "bottomright",
          legend = c(
              sprintf("AUC = %.3f", auc(roc_obj)),
              sprintf("F1 = %.3f", f1),
              sprintf("Accuracy = %.3f", accuracy),
              sprintf("Precision = %.3f", precision),
              sprintf("Recall = %.3f", recall)
          )
      )
}

glm_predict_func <- function(model, newdata) predict(model, newdata, type = "response")
rpart_predict_func = function(model, newdata) predict(model, newdata, type = "prob")[, "yes"]
```

Let us define weights as inverse of classes' rates. that way both classes' weights sum up to the same number.
```{r}
weights <- as.double(data$y)
pos_rate <- mean(data$y == 'yes')
weights[data$y == 'yes'] <- 1 / pos_rate 
weights[data$y == 'no']  <- 1 / (1 - pos_rate)
```

## Baseline Models

We will create dummy estimator, that predicts the most frequent class, logistic regression and decision tree.

### Dummy Estimator

```{r}
dummy_model <- function(fml, data, ...) {
    data$y
}

dummy_predict <- function(train_labels, test_data) {
    majority_class <- ifelse(names(sort(table(train_labels), decreasing = TRUE))[1] == 'no', 0, 1)
  rep(majority_class, nrow(test_data))
}
cv_evaluate_model(dummy_model, dummy_predict)
```

### Logistic Regression
```{r}
cv_evaluate_model(glm, glm_predict_func, family=binomial)
```

```{r}
cv_evaluate_model(glm, glm_predict_func, family=binomial, weights=weights)
```

### Decision Tree Classifier
```{r}
cv_evaluate_model(rpart, rpart_predict_func, maxdepth=3, cp=0)
```

```{r}
cv_evaluate_model(rpart, rpart_predict_func, weights = weights, maxdepth=3, cp=0)
```

## Feature Selection

### Lasso
```{r}
X <- model.matrix(y ~ ., data = data)[, -1]  # remove intercept

y <- data$y
y <- ifelse(y == "yes", 1, 0)

model <- cv.glmnet(X, y, family = "binomial", alpha = 1)
coefs <- coef(model, s = 'lambda.1se')  # or use s = "lambda.min" if using cv.glmnet
coefs <- as.matrix(coefs)

# Get variable names with non-zero coefficients
non_zero_vars <- rownames(coefs)[which(coefs != 0)]
non_zero_vars <- setdiff(non_zero_vars, "(Intercept)")


print(coefs)
non_zero_vars
```

```{r}
lasso_data <- data[,c("y", "job", "marital", "education", "housing", "loan", "contact", "month", "duration", "poutcome")]
cv_evaluate_model(glm, glm_predict_func, family=binomial, arg_data = lasso_data)
```

```{r}
cv_evaluate_model(glm, glm_predict_func, family=binomial, arg_data = lasso_data, weights = weights)
```

```{r}
build_formula <- function(arg_data, target = 'y', exclude = NULL) {
  predictors <- setdiff(names(arg_data), c(target, exclude))
  as.formula(paste(target, "~", paste(predictors, collapse = " + ")))
}

X <- model.matrix(build_formula(data, exclude = 'month'), data = data)[, -1]  # remove intercept

y <- data$y
y <- ifelse(y == "yes", 1, 0)

model <- cv.glmnet(X, y, family = "binomial", alpha = 1)
coefs <- coef(model, s = 'lambda.1se')  # or use s = "lambda.min" if using cv.glmnet
coefs <- as.matrix(coefs)

# Get variable names with non-zero coefficients
non_zero_vars <- rownames(coefs)[which(coefs != 0)]
non_zero_vars <- setdiff(non_zero_vars, "(Intercept)")


print(coefs)
non_zero_vars
```

```{r}
lasso_data <- data[,c("y", "job", "marital", "housing", "loan", "contact", "duration", "poutcome")]
cv_evaluate_model(glm, glm_predict_func, family=binomial, arg_data = lasso_data)
```

```{r}
cv_evaluate_model(glm, glm_predict_func, family=binomial, arg_data = lasso_data, weights = weights)
```

### Step with AIC heuristics

```{r}
step(glm(y ~ ., data, family='binomial'), direction='both', trace=0)
```

```{r}
step_data <- data[,c("y", "marital", "education", "housing", "loan", "contact", "day", "month", "duration", "campaign", "poutcome")]
cv_evaluate_model(glm, glm_predict_func, family=binomial, arg_data = step_data)
```

```{r, warning=FALSE}
step(glm(y ~ ., data, family='binomial', weights = weights), direction='both', trace=0)
```

```{r}
step_data <- data[,c("y", "marital", "education", "housing", "loan", "contact", "day", "month", "duration", "campaign", "poutcome")]
cv_evaluate_model(glm, glm_predict_func, family=binomial, arg_data = lasso_data)
```
```{r}
step(glm(y ~ ., data[,setdiff(names(data), c('month'))], family='binomial'), direction='both', trace=0)
```

```{r}
step_data <- data[,c("y", "job", "marital", "education", "housing", "loan", "contact", "duration", "campaign", "poutcome")]
cv_evaluate_model(glm, glm_predict_func, family=binomial, arg_data = step_data)
```

```{r, warning=FALSE}
step(glm(y ~ ., data[,setdiff(names(data), c('month'))], family='binomial', weights = weights), direction='both', trace=0)
```

```{r}
step_data <- data[,c("y", "job", "marital", "education", "housing", "loan", "contact", "duration", "campaign", "poutcome")]
cv_evaluate_model(glm, glm_predict_func, family=binomial, weights=weights, arg_data = step_data)
```


### Job aggregation
```{r}
data.specific_job <- data
data.specific_job[!(data$job == 'blue-collar' | data$job == 'retired'),"job"] <- "unknown"
data.specific_job
```

```{r}
X <- model.matrix(build_formula(data.specific_job, exclude = 'month'), data = data)[, -1]  # remove intercept

y <- data$y
y <- ifelse(y == "yes", 1, 0)

model <- cv.glmnet(X, y, family = "binomial", alpha = 1)
coefs <- coef(model, s = 'lambda.1se')  # or use s = "lambda.min" if using cv.glmnet
coefs <- as.matrix(coefs)

# Get variable names with non-zero coefficients
non_zero_vars <- rownames(coefs)[which(coefs != 0)]
non_zero_vars <- setdiff(non_zero_vars, "(Intercept)")


print(coefs)
non_zero_vars
```

```{r}
lasso_data <- data.specific_job[,c("y", "job", "marital", "housing", "loan", "contact", "duration", "poutcome")]
cv_evaluate_model(glm, glm_predict_func, family=binomial, arg_data = lasso_data)
```
```{r}
step(glm(y ~ ., data=data.specific_job[,setdiff(names(data), c('month'))] , family = binomial), direction = 'both', trace=0)
```

```{r}
step_data <- data.specific_job[,c("y", "job", "marital", "education", "housing", "loan", "contact", "duration", "campaign", "poutcome")]
cv_evaluate_model(glm, glm_predict_func, family=binomial, arg_data = step_data)
```

## Final model
lasso, no month, not weighted, joint jobs
```{r}
final_model <- glm(y ~ job + marital + housing + loan + contact + poutcome + duration, data=data.specific_job, family=binomial)
summary(final_model)
```

## Assumption Verification

### Linearity of Continuous Variable Logits 

```{r}
data.specific_job.numeric <- data.specific_job
data.specific_job.numeric$y <- ifelse(data.specific_job$y == "yes", 1, 0)
car::boxTidwell(y ~ duration, data = data.specific_job.numeric)
```

```{r}
lambda = 0.79334
data.specific_job.numeric$duration_transformed <- data.specific_job$duration ^ lambda
car::boxTidwell(y ~ duration_transformed, data = data.specific_job.numeric)
```

```{r}
final_model <- glm(y ~ job + marital + housing + loan + contact + poutcome + I(duration ^ lambda), data=data.specific_job, family=binomial)
summary(final_model)
```

### Outliers

```{r}
plot(final_model, which=4)
plot(final_model, which=5)
```

```{r}
final_model <- glm(y ~ job + marital + housing + loan + contact + poutcome + I(duration ^ lambda), data=data.specific_job[-569,], family=binomial)
summary(final_model)
```

```{r}
plot(final_model, which=4)
plot(final_model, which=5)
```
### Multicollinearity

```{r}
car::vif(final_model)
```

## Final Decision Tree

```{r}
data.lasso_specific_job <- data.specific_job[,c("y", "job", "marital", "housing", "loan", "contact", "duration", "poutcome")]
cv_evaluate_model(rpart, rpart_predict_func, arg_data = data.lasso_specific_job)
```

```{r}
tree_model <- rpart(y ~ ., data.lasso_specific_job, maxdepth=3, cp=0)
rpart.plot(tree_model)
```

## Wrap up

```{r}
saveRDS(final_model, file=here::here("models", "log_regression.rds"))
saveRDS(tree_model,  file=here::here("models", "decision_tree.rds"))
```