---
title: "04_profiling"
author: "Maksym Khavil"
date: "2025-07-23"
output: html_document
---

## Setup 

Let us load the data and libraries.
```{r}
library(stats)
library(rpart)
library(rpart.plot)
library(corrplot)
library(car)
library(vioplot)
library(effsize)

source(here::here("R", "constants.R"))
source(here::here("R", "load_data.R"))

data <- load_bank_data()
logistic_regression <- readRDS(here::here("models", "log_regression.rds"))
decision_tree <- readRDS(here::here("models", "decision_tree.rds"))
head(data)
```

```{r}
data.joint_job <- data
data.joint_job[!(data$job == 'blue-collar' | data$job == 'retired'),"job"] <- "unknown"
data.pos <- data[data$y == 'yes',]
data.joint_job.pos <- data.joint_job[data.joint_job$y == 'yes',]

summary(logistic_regression)
rpart.plot(decision_tree)
```

### Data separatation
```{r}
plot_count_bars <- function(col, arg_data, ...) {
    pos_labeled = arg_data[arg_data$y == "yes",]
    neg_labeled = arg_data[arg_data$y == "no",]
    pos <- aggregate(x = list(count = pos_labeled$y), by = list(column = pos_labeled[,col]), FUN = length)
    neg <- aggregate(x = list(count = neg_labeled$y), by = list(column = neg_labeled[,col]), FUN = length)
    data_matrix <- matrix(
        c( pos$count,  neg$count), nrow = 2, byrow = TRUE
    )
    colnames(data_matrix) <- pos$column
    rownames(data_matrix) <- c("Pos. Cnt", "Neg. Cnt")
    barplot(
        data_matrix,
        beside = TRUE,
        legend = rownames(data_matrix),
        main = sprintf("Class Counts X %s", col),
        xlab = col, ylab = "Class Counts",
        ...
    )
    grid()
}

plot_rate_bars <- function(col, arg_data, ...) {
    pos_labeled = arg_data[arg_data$y == "yes",]
    neg_labeled = arg_data[arg_data$y == "no",]
    par(mar=c(5, 4, 6, 2))
    total_cnt <- aggregate(x = list(count = arg_data$y), by = list(column = arg_data[,col]), FUN = length)$count
    pos <- aggregate(x = list(count = pos_labeled$y), by = list(column = pos_labeled[,col]), FUN = length)
    neg <- aggregate(x = list(count = neg_labeled$y), by = list(column = neg_labeled[,col]), FUN = length)
    data_matrix <- matrix(
        c( pos$count / total_cnt,  neg$count / total_cnt), nrow = 2, byrow = TRUE
    )
    colnames(data_matrix) <- pos$column
    rownames(data_matrix) <- c("Pos. Rate", "Neg. Rate")
    barplot(
        data_matrix,
        main = sprintf("Class Rates X %s", col),
        ylim=c(0,1),
        xlab = col, ylab = "Class Rates",
        ...
    )
    grid()
    legend(
        "topright",
        legend = rownames(data_matrix),
        xpd = T,
        fill=c("black", "grey"),
        inset=c(0, -0.2)
    )
}
```

```{r}
data.joint_job$score <- predict(logistic_regression, newdata=data.joint_job, type='response')
data.joint_job$segment <- cut(data.joint_job$score, breaks = c(-Inf, 0.3, 0.7, Inf), labels = c("Low", "Medium", "High"), ordered_result = T)

data.joint_job.pos$score <- predict(logistic_regression, newdata = data.joint_job.pos, type='response')
data.joint_job.pos$segment <- cut(data.joint_job.pos$score, breaks = c(-Inf, 0.3, 0.7, Inf), labels = c("Low", "Medium", "High"), ordered_result = T)

plot_count_bars("segment", data.joint_job)
plot_rate_bars( "segment", data.joint_job)
```

```{r}
data.joint_job.low    <- data.joint_job[data.joint_job$segment == "Low",]
data.joint_job.medium <- data.joint_job[data.joint_job$segment == "Medium",]
data.joint_job.high   <- data.joint_job[data.joint_job$segment == "High",]
```

## Data analysis

```{r}
for(col in colnames(data.joint_job[sapply(data.joint_job, is.factor)])) {
    if(col == "segment" || col == "y"){ next}
    ct = table(data.joint_job[,col], data.joint_job$segment)
    par(mar=c(3,4,1,2))
    mosaicplot(
        ct, ylab="Segment", xlab=col,
        main=sprintf("Segment X %s", col),
        las = 2
    )
}
```

```{r}
ylim=c(0, 1)
for(col in colnames(data.joint_job[sapply(data.joint_job, is.numeric)])) {
    if(col == "score") {next}
    fml <- as.formula(paste(col, " ~ segment"))
    par(mfrow=c(1, 2))
    boxplot(fml, data.joint_job, col=COLORS[col])
    vioplot(fml, data.joint_job, col=COLORS[col])
}
```

## Creating Profiles

housing = no
loan = yes

job = "retired", "student"

poutcome = success
pdays > 0


```{r}
num_y <- ifelse(data$y == "yes", 1, 0)
cohen.h <- function(p1, p2) {
  2 * asin(sqrt(p1)) - 2 * asin(sqrt(p2))
}
analyse_profile <- function(profile_bool, baseline.data = num_y) {
    num_profile <- ifelse(baseline.data[profile_bool], 1, 0)
    num_comp_profile <- ifelse(baseline.data[!profile_bool], 1, 0)
    
    print("--------------------------------------------------", quote=F)
    print(sprintf("Coverage:        %d samples = %.1f%% of dataset", length(num_profile), 100 * length(num_profile) / length(baseline.data)), quote=F)
    print(sprintf("Positives:       %d samples = %.1f%% of profile", sum(num_profile), 100 * mean(num_profile)), quote=F)
    print("--------------------------------------------------", quote=F)
    if (length(num_profile) < 1 || sum(num_profile) < 1) return()
    
    
    p_profile <- mean(num_profile)
    p_baseline <- mean(baseline.data)
    
    prop_pvalue <- prop.test(x = c(sum(num_profile), sum(num_comp_profile)), n = c(length(num_profile), length(num_comp_profile)), alternative = "less")$p.value
    binom_pvalue <- binom.test(sum(num_profile), length(num_profile), p=p_baseline, alternative = "less")$p.value # estimate
    print(sprintf("Proportion test p-value against profile's compliment:     %.3f", prop_pvalue), quote=F)
    print(sprintf("Exact Binomial test p-value:                              %.3f", binom_pvalue), quote=F)
    
    print("--------------------------------------------------", quote=F)
    cohen_h <- cohen.h(p_profile, p_baseline)
    risk_difference <- p_profile - p_baseline
    odds_ratio <- p_profile * (1 - p_profile) / (p_baseline * (1 - p_baseline))
    
    
    print(sprintf("Cohen's h value: %.3f", cohen_h), quote=F)
    print(sprintf("Risk difference: %.3f", risk_difference), quote=F)
    print(sprintf("Odds ratio:      %.3f", odds_ratio), quote=F)
    print("--------------------------------------------------", quote=F)
    print("", quote=F)
}
```

### Univariate
```{r}
analyse_profile(data$housing == "no")
```


```{r}
analyse_profile(data$loan == "yes")
```

```{r}
print("Job = student")
analyse_profile(data$job == "student")
print("Job = retired")
analyse_profile(data$job == "retired")
```

```{r}
print("Previous outcome = success")
analyse_profile(data$poutcome == "success")
print("Previous outcome = other")
analyse_profile(data$poutcome == "other")
print("Previous outcome = other or success")
analyse_profile((data$poutcome == "success") | (data$poutcome == "other"))
```

```{r}
analyse_profile(data$pdays > 0)
```

```{r}
print("Balance greater than 0")
analyse_profile(data$balance > 0)
print("Balance greater than 100")
analyse_profile(data$balance > 100)
print("Balance greater than 500")
analyse_profile(data$balance > 500)
print("Balance greater than 1000")
analyse_profile(data$balance > 1000)
```

### Bivariate

```{r}
print("Previous outcome = success & housing = no")
analyse_profile((data$poutcome == "success") & (data$housing == "no"))
print("Previous outcome = success & loan = yes")
analyse_profile((data$poutcome == "success") & (data$loan == "yes"))
```

```{r}
print("Previous outcome = other & housing = no")
analyse_profile((data$poutcome == "other") & (data$housing == "no"))
print("Previous outcome = other & loan = yes")
analyse_profile((data$poutcome == "other") & (data$loan == "yes"))
```

```{r}
print("Previous outcome = (other or success) & housing = no")
analyse_profile((data$poutcome %in% c("other", "success")) & (data$housing == "no"))
print("Previous outcome = (other or success) & loan = yes")
analyse_profile((data$poutcome %in% c("other", "success")) & (data$loan == "yes"))
```

```{r}
print("Job = retired & housing = no")
analyse_profile((data$job == "retired") & (data$housing == "no"))
print("Job = retired & loan = yes")
analyse_profile((data$job == "retired") & (data$loan == "yes"))
```

```{r}
print("Job = student & housing = no")
analyse_profile((data$job == "student") & (data$housing == "no"))
print("Job = student & loan = yes")
analyse_profile((data$job == "student") & (data$loan == "yes"))
```

```{r}
analyse_profile((data$pdays > 0) & (data$previous %in% c("other", "success")))
```

```{r}
analyse_profile((data$loan == "yes") | (data$housing == "no"))
```
```{r}
analyse_profile((data$job == "retired") & (data$balance > 0))
```

```{r}
analyse_profile((data$housing == "no") & (data$pdays > 0))
```

## Profiles

Hosing = no and pdays > 0 (have being previously contacted and have capacity to subscribe to a deposit)

job = retired and loan = yes (retired people who have capacity and motivation to grow their fortune)

## Test on full dataset

```{r}
full_data <- load_bank_data("bank-full.csv")
```

```{r}
num_y_full <- ifelse(full_data$y == "yes", 1, 0)
analyse_profile((full_data$housing == "no") & (full_data$pdays > 0), baseline.data = num_y_full)
```

```{r}
analyse_profile((full_data$job == "retired") & (full_data$balance > 0), baseline.data = num_y_full)
```

Profile intersection
```{r}
analyse_profile((full_data$job == "retired") & (full_data$balance > 0) & (full_data$housing == "no") & (full_data$pdays > 0), baseline.data = num_y_full)
```

## Conclusion


